---
title: "Linkin Park’s Musical Transition"
subtitle: "From Acoustic-Influenced Rock to Electronically Produced Tracks"
author: 
  - Ziheng Zhong
  - YingKe He
thanks: "Code and data are available at: https://github.com/iJustinn/Musical_Analysis.git"
date: today
date-format: long
abstract: "This paper analyzes the evolution of Linkin Park’s music by examining key audio features such as acousticness, danceability, and energy using data from the Spotify API. The study focuses on the band’s most popular albums, highlighting their shift from acoustic-influenced rock to more electronically produced and rhythmically accessible tracks over time. Despite these changes, Linkin Park consistently maintained high energy in their music. The results provide understandings into the band’s adaptability and their ability to evolve while retaining core elements of their signature sound."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# Library Setup
library(here)
library(readr)
library(dplyr)
library(usethis)
library(magrittr)
library(spotifyr)
library(tidyverse)
library(lubridate)
library(kableExtra)

# Load the data
raw_data <- readRDS(here("data", "raw_data", "linkin_park_data.rds"))
chart_1_data <- readRDS(here("data", "analysis_data", "clean_data_1.rds"))
```



# Introduction

Linkin Park is a globally renowned band known for its innovative fusion of rock, electronic, and alternative music. Since their debut album Hybrid Theory in 2000, the band has consistently evolved its sound, blending genres and experimenting with new styles. This paper seeks to analyze the sonic evolution of Linkin Park’s music through key audio features such as acousticness, danceability, and energy, using data sourced from the Spotify API. By examining these attributes across their discography, the study aims to uncover trends in their musical style and provide insights into how the band has adapted to changes in the music industry over time.

Focusing on their most popular albums, including Hybrid Theory, Meteora, and One More Light, this paper explores how Linkin Park shifted from the acoustic-influenced, high-energy rock tracks of their early years to more rhythmically accessible and electronically produced music in later albums. Through this analysis, the paper not only captures the band’s stylistic transitions but also highlights how they maintained core elements that have defined their music. By combining quantitative audio data with musical trends, this research offers a comprehensive view of Linkin Park’s evolving sound over two decades.

The structure of this paper is organized as follows: @sec-data presents the data sources, methodologies, and the visualization of the results. The findings are then discussed in @sec-dis, and additional detailed information is provided in the appendix at @sec-app.



# Data {#sec-data}
This project is motivated and guided by Rohan Alexander and his book [@citeTbook]. Data used in this paper was downloaded, cleaned and analyzed with the programming language `R` [@citeR]. Also with the support of additional packages in R: `dplyr` [@citeDplyr], `usethis` [@citeUsethis], `magrittr` [@citeMagrittr], `spotifyr` [@citeSpotifyr], `tidyverse` [@citeTidyverse], `lubridate` [@citeLubridate], `here` [@citeHere] and `readr` [@citeReadr].

## Source

In this paper, data regarding the band Linkin Park was fetched from the Spotify API [@citeSpotify]. The Spotify API provides comprehensive music data, including track details, album information, artist popularity metrics, and more. By querying this API, we were able to fetch key information about Linkin Park, such as track popularity, album releases, and other metadata associated with their discography. The API's ability to deliver reliable data ensures that the information used in this analysis reflects the band's ongoing listener engagement and global presence.

Method of how data was processed can be found in @sec-app.

## Measurement
Spotify doesn't release detailed methodologies on how acousticness, energy, and danceability are measured, the below are are potential approaches to how these features may be quantified:

### Acousticness
This feature likely evaluates the presence of non-electronic instrumentation or sounds within a track. With acoustic music relying on instruments like guitars or pianos. The measurement could be based on the absence of electronic frequencies or specific sound wave patterns commonly produced by acoustic instruments. 

### Energy
Energy measures the intensity and activity of a track. It could be calculated through several factors, such as tempo, perceived loudness, and how much the volume changes within the track. So tracks with fast tempos, prominent beats, and higher average volumes might be assigned higher energy values.

### Danceability
Danceability likely accounts for a track's rhythm stability, tempo, and overall beat strength. A song's suitability for dancing might be measured by its consistent and predictable beat, minimal rhythmic complexity, and tempo that falls within a range conducive to most types of dance music. This could be evaluate by analyzing the regularity and strength of beats or the prominence of rhythmic elements.

## Results

```{r}
#| echo: false
#| eval: true
#| label: fig-energy
#| fig-cap: "Distribution of Energy of Linkin Park Songs Over Time (2002-2023)"
#| warning: false
#| message: false
#| fig-height: 6
#| fig-width: 7

chart_1_data %>%
  ggplot(aes(x = album_release_date, y = energy, color = album_name)) +
  geom_line(stat = "summary", fun = "mean", size = 1.2) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(
    x = "Album Release Date",
    y = "Energy",
    color = "Album"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.key.width = unit(1.5, "cm"),
    legend.box = "horizontal"
  ) +
  guides(color = guide_legend(nrow = 5, byrow = TRUE))

```

@fig-energy illustrates the acousticness of Linkin Park’s songs from their various albums over time. Acousticness, which measures how acoustic a track is, varies significantly across the band’s discography. Earlier albums, particularly around the early 2000s, have a broad range of acousticness, with some tracks reaching nearly 1.00 (high acoustic quality), while most remain closer to the 0.0 end of the scale. As time progresses, the acousticness generally decreases, indicating a preference for more electronically produced sounds in later albums. Notably, albums like "A Thousand Suns" and "The Hunting Party" show songs with relatively low acousticness values, emphasizing the band's shift toward more energetic and electronic compositions.

```{r}
#| echo: false
#| eval: true
#| label: fig-dance
#| fig-cap: "Distribtion of Danceability of Linkin Park Songs Over Time (2002-2023)"
#| warning: false
#| message: false
#| fig-height: 6
#| fig-width: 7

chart_1_data %>%
  ggplot(aes(x = album_release_date, y = danceability, color = album_name)) +
  geom_line(stat = "summary", fun = "mean", size = 1.2) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(
    x = "Album Release Date",
    y = "Danceability",
    color = "Album"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.key.width = unit(1.5, "cm"),
    legend.box = "horizontal"
  ) +
  guides(color = guide_legend(nrow = 5, byrow = TRUE))

```

@fig-dance showcases the danceability scores of Linkin Park’s tracks, which represent how suitable the songs are for dancing based on tempo, rhythm stability, and beat strength. Across their early albums, danceability shows moderate variation, with several songs scoring between 0.4 and 0.6. As the timeline advances, there’s a noticeable increase in danceability, especially in albums like "One More Light," released post-2010, where multiple tracks score higher than previous ones. This indicates a transition in the band's music style, incorporating more rhythmically accessible and upbeat elements, perhaps in line with the trends of more modern music.

```{r}
#| echo: false
#| eval: true
#| label: fig-acoustic
#| fig-cap: "Distribtion of Acousticness of Linkin Park Songs Over Time (2002-2023)"
#| warning: false
#| message: false
#| fig-height: 6
#| fig-width: 7

chart_1_data %>%
  ggplot(aes(x = album_release_date, y = acousticness, color = album_name)) +
  geom_line(stat = "summary", fun = "mean", size = 1.2) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(
    x = "Album Release Date",
    y = "Acousticness",
    color = "Album"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.key.width = unit(1.5, "cm"),
    legend.box = "horizontal"
  ) +
  guides(color = guide_legend(nrow = 5, byrow = TRUE))

```

@fig-acoustic visualizes the energy levels of Linkin Park songs, with energy representing the intensity and activity of the tracks. The data shows that the band maintained consistently high-energy tracks across their entire discography. Songs from early albums like "Hybrid Theory" and "Meteora" exhibit high energy levels, typically above 0.75, contributing to the aggressive, dynamic sound the band is known for. Even in their later works, such as "The Hunting Party" and "One More Light," energy levels remain elevated, reflecting Linkin Park's continuous emphasis on powerful, fast-paced music despite stylistic changes over the years.

# Discussion {#sec-dis}

## The transition

The results from the acousticness, danceability, and energy charts highlight Linkin Park’s evolving sound over time. Early albums like Hybrid Theory and Meteora exhibit a wide range of acousticness, reflecting their experimental blend of rock and electronic elements. As the band's music matured, particularly from A Thousand Suns onward, their reliance on acoustic elements decreased, aligning more with electronic production styles. This shift is further supported by the increasing danceability in later albums like One More Light, suggesting the band's exploration into more rhythmically accessible and pop-influenced tracks. Throughout their career, Linkin Park maintained a high-energy sound, which remained a hallmark of their music, regardless of shifts in style. These results underscore the band’s ability to adapt and evolve while still retaining core elements that define their musical identity.

## Weaknesses and next steps

One limitation of this analysis is its reliance on a few quantitative measures (acousticness, danceability, energy) to represent the complexity of Linkin Park's music. These metrics, while useful, do not fully capture other important aspects such as lyrical content, emotional intensity, and production intricacies that contribute to the overall character of their music. Additionally, the data only focuses on a subset of their albums, potentially overlooking valuable insights from live performances, collaborations, and lesser-known releases. Future projects should consider a more comprehensive dataset, including these factors, and apply more sophisticated methods like sentiment analysis or lyrical theme extraction to provide a deeper, multi-dimensional understanding of the band’s music evolution.

\newpage



\appendix

# Appendix {#sec-app}

```{r}
#| echo: false
#| eval: true
#| label: tbl-raw
#| tbl-cap: "Raw Linkin Park Spotify Data"
#| warning: false
#| message: false

selected_raw_data <- raw_data[, c("album_id", "danceability", "energy", "acousticness")]
selected_raw_data[, 2:4] <- round(selected_raw_data[, 2:4], 2)

kbl(selected_raw_data[1:6, ]) %>%
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed"))

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-clean
#| tbl-cap: "Processed Linkin Park Spotify Data"
#| warning: false
#| message: false

selected_chart_data <- chart_1_data[, c("album_id", "danceability", "energy", "acousticness")]
selected_chart_data[, 2:4] <- round(selected_chart_data[, 2:4], 2)

kbl(selected_chart_data[1:6, ]) %>%
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed"))

```

The data processing starts by narrowing the dataset to focus on the top 10 most popular albums by Linkin Park. A list called popular_albums is created, containing the names of these albums, including well-known releases like "Hybrid Theory," "Meteora (Bonus Edition)," and "Minutes to Midnight." This list helps to filter the dataset, ensuring that only the most significant and impactful albums are retained for analysis. These changes can be found by comparing @tbl-raw and @tbl-clean.

Next, the code handles potential inconsistencies in the `album_release_date` field by ensuring that all dates are properly formatted. The `mutate()` function is used in combination with the `ymd()` function, which attempts to parse the date strings into a standard year-month-day format. To avoid disruptive errors caused by parsing issues, `suppressWarnings()` is used, which suppresses any warning messages if the date format is invalid or unparseable. This step is essential because incorrect or inconsistent date formats could skew analysis results.

After ensuring that the `album_release_date` is correctly formatted, the data is further cleaned by filtering out any rows where the date parsing failed. This is done by checking for missing or `NA` values in the `album_release_date` field. Rows with `NA` values are excluded from the dataset using the `filter()` function. This final step guarantees that only the records with valid release dates are included, ensuring the reliability of subsequent analyses.

\newpage


# References


